Blockchain System

### 2020-12-22 papers/execution-model/model.tex
further simplification to local node

\begin{definition}%[Operation]
  \emph{Operations} are defined by the following grammar:
  \begin{align*}
    \OP &::= \TRANSFER[\PARAMETER]\NTEZ\PUK\ADDR\MTEZ
    \\&\mid \ORIGINATE\NTEZ\PUK\CODE\INIT\MTEZ
  \end{align*}
  where \NTEZ\ and \MTEZ\ are amount of Tez, \ADDR\ is either a public key for an implicit account  or a public hash for a smart contract,  \PARAMETER\ is an argument passed to the smart contract's script, which is emtry if it is a transfer to an implicit account, \CODE\ is a script of the smart contract, \INIT\ is an initial value of the contract's storage.
\end{definition}

@@

 \begin{definition}%[Pending operation]
An \emph{operation} is represented by a tuple  $\langle  \OP, \TIME, \STATUS
\rangle $, where $\OP$ is an operation, $\TIME$ is the time when the
operation was injected, and $\STATUS$ is either $\STATUSPENDING$,
$\STATUSINCLUDING\ \TIME'$, or $\STATUSTIMEOUT$.  
\end{definition}
\begin{definition}
  The pool of operations is defined as a partial mapping from
  operation hashes $\OPH$:
  \begin{align*}
    \PENDING &= \TOPH \partialto \TOP\times \TTIME \times \TSTATUS \\
    \TSTATUS&= (\STATUSPENDING + \STATUSINCLUDED\ \TTIME + \STATUSTIMEOUT)
  \end{align*}
\end{definition}

@@

\begin{definition}%[Blockchain system]
A \emph{blockchain configuration} is a pair
$ \NODE \| \BLOCKCHAIN$ where
\fxnote[inline]{Should the model contain multiple nodes?}
\begin{enumerate}
\item $\NODE = [\EXPRS, \ACCOUNTS, \OPERATIONS]$ is the state of a node, and
\item $\BLOCKCHAIN = [\PENDING, \MANAGERS, \CONTRACTORS,
  \TIME]$ is the state of a blockchain  
  such that
  \begin{itemize}
  \item
    $\forall c \in \ACCOUNTS \implies \exists k\in \MANAGERS, k.\PUK =
    c.\PUK$;
  \item if $\PENDING (\OPH) = \Angle{\OP,\TIME, \STATUSINCLUDED\ \TIME'}$,
    then $\TIME' > \TIME$
  \end{itemize}
\end{enumerate}
\end{definition}

\clearpage{}
\section{Rules}

\begin{mathpar}
  \inferrule[Node-Eval]
  {
    \EXPR \ExprTrans \EXPR'
  }{
    [\EC\EXPR :: \EXPRS, \ACCOUNTS, \OPERATIONS] \NodeTrans{}
    [\EC{\EXPR'} :: \EXPRS, \ACCOUNTS, \OPERATIONS]
  }
\end{mathpar}

\begin{mathpar}
  \inferrule[Node-Op]
    {
    \Angle{\PAK,\OP.\PUK} \in \ACCOUNTS
  }{
    [\EC\OP :: \EXPRS, \ACCOUNTS, \OPERATIONS] \NodeTrans{}
    [\EXPRS, \ACCOUNTS, (\EC{},\OP):: \OPERATIONS]
  }
\end{mathpar}

@@

\begin{mathpar}
  \inferrule{
    \CHECKBAL (\MANAGERS, \PUK, \NTEZ, \MTEZ) \\
    \CHECKCOU (\MANAGERS, \PUK) \\
    \CHECKPUB (\MANAGERS, \PUK') \\
    \CHECKGAS (\PUK, \PUK', \NTEZ, \MTEZ) \\
    \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
    \OPH = \GENERATEOPH (\PUK, \PUK', \NTEZ, \MTEZ, \TIME)
  }{
   { [\ACCOUNTS, \OP :: \OPERATIONS] \| [\PENDING,
     \ACCEPTED, \MANAGERS, \CONTRACTORS, \TIME] } \SystemTrans \\
   { [\EXPRS, \ACCOUNTS, \OPERATIONS] \|
     [ \Angle{\OP,\OPH, \TIME}
     ::\PENDING,
     \ACCEPTED,
     \UPDATECOU(\MANAGERS, \PUK, \TRUE),
     \CONTRACTORS,
     \TIME]
   }
  }
\end{mathpar}
~\\
~\\
\fi
Rule 1 [rejected of counter]:
\begin{mathpar}
  \inferrule{
    %\NEG\ \CHECKCOU (\MANAGERS, \PUK)\\
    %\OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
    \NEG\ \CHECKCOU (\MANAGERS, \PUK) \\
    \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
  }{
   { [\EXPRS, \ACCOUNTS, (\EC{},\OP) :: \OPERATIONS] \| [\PENDING,
     \ACCEPTED, \MANAGERS, \CONTRACTORS, \TIME] } \SystemTrans
   { [\EXPRS, \ACCOUNTS, \OPERATIONS] \|
     [ \PENDING,
     \ACCEPTED,
     \MANAGERS,
     \CONTRACTORS,
     \TIME]
   }
  }
\end{mathpar}
~\\
~\\
Rule 2 [rejected of balance]:
\begin{mathpar}
  \inferrule{
    \NEG\ \CHECKBAL (\MANAGERS, \PUK, \NTEZ, \MTEZ) \\
    \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
  }{
   { [\EXPRS, \ACCOUNTS, (\EC{},\OP) :: \OPERATIONS] \| [\PENDING,
     \ACCEPTED, \MANAGERS, \CONTRACTORS, \TIME] } \SystemTrans
   { [\EXPRS, \ACCOUNTS, \OPERATIONS] \|
     [ \PENDING,
     \ACCEPTED,
     \MANAGERS,
     \CONTRACTORS,
     \TIME]
   }
  }
\end{mathpar}
~\\
~\\
Rule 3 [rejected of public key]:
\begin{mathpar}
  \inferrule{
    \NEG\ \CHECKPUB (\MANAGERS, \PUK') \\
    \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
  }{
   { [\EXPRS, \ACCOUNTS, (\EC{},\OP) :: \OPERATIONS] \| [\PENDING,
     \ACCEPTED, \MANAGERS, \CONTRACTORS, \TIME] } \SystemTrans
   { [\EXPRS, \ACCOUNTS, \OPERATIONS] \|
     [ \PENDING,
     \ACCEPTED,
     \MANAGERS,
     \CONTRACTORS,
     \TIME]
   }
  }
\end{mathpar}
~\\
~\\
Rule 4 [rejected of fee]:
\begin{mathpar}
  \inferrule{
    \NEG\ \CHECKGAS (\PUK, \PUK', \NTEZ, \MTEZ) \\
    \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
  }{
   { [\EXPRS, \ACCOUNTS, (\EC{},\OP) :: \OPERATIONS] \| [\PENDING,
     \ACCEPTED, \MANAGERS, \CONTRACTORS, \TIME] } \SystemTrans
   { [\EXPRS, \ACCOUNTS, \OPERATIONS] \|
     [ \PENDING,
     \ACCEPTED,
     \MANAGERS,
     \CONTRACTORS,
     \TIME]
   }
  }
\end{mathpar}
\iffalse 
Rule 6 [included]:
\begin{mathpar}
  \inferrule[Block-Accept]{
    \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
    \TIME'-\TIME < 60
  }{
    {[\Angle{\OP,\OPH, \TIME} :: \PENDING, \ACCEPTED, \MANAGERS,
      \CONTRACTORS, \TIME']}
    \BlockTrans  \\
    {[\PENDING, \Angle{\OP,\OPH, \TIME,\TIME'} :: \ACCEPTED,
      \UPDATESUCC(\MANAGERS, \PUK, \PUK', \NTEZ, \MTEZ),
      \CONTRACTORS, \TIME'+1]}
  }
\end{mathpar}
Rule 7 [timeout]: (applies to both, implicit transfers and contract invocations)
\begin{mathpar}
  \inferrule[Block-Timeout]{
    \OP = \TRANSFER[\STRING]\NTEZ\PUK\ADDR\MTEZ \\
    \TIME'-\TIME \ge 60
  }{
    {[\Angle{\OP,\OPH, \TIME} :: \PENDING, \ACCEPTED, \MANAGERS,
      \CONTRACTORS, \TIME']}
    \BlockTrans
    {[\PENDING, \ACCEPTED, \UPDATECOU(\MANAGERS, \PUK, \FALSE),
      \CONTRACTORS, \TIME']}
  }
\end{mathpar}
\fi
\subsection{Smart Contracts}
A. Originate
~\\
~\\
Rule 1 [injected]:

>>>>

\begin{definition}%[Operation]
  \emph{Operations} are defined by the following grammar:
  \begin{align*}
    \OP &::= \TRANSFER[\PARAMETER]\NTEZ\PUK\ADDR\MTEZ
    \\&\mid \ORIGINATE\NTEZ\PUK\CODE\INIT\MTEZ
  \end{align*}
  where \NTEZ\ and \MTEZ\ are tokens, \ADDR\ is either a public
  key for an implicit account  or a public hash for a smart contract,
  \PARAMETER\ is an argument passed to the smart contract's script,
  which is empty if it is a transfer to an implicit account, \CODE\ is
  a script of the smart contract, \INIT\ is an initial value of the
  contract's storage. 
\end{definition}

@@

 \begin{definition}%[Pending operation]
An \emph{operation} is represented by a tuple  $\langle  \OP, \TIME, \STATUS
\rangle $, where $\OP$ is an operation, $\TIME$ is the time when the
operation was injected, and $\STATUS$ is either $\STATUSPENDING$,
$\STATUSINCLUDING\ \TIME'$, or $\STATUSTIMEOUT$.  
\end{definition}
\begin{definition}
  The pool of operations is defined as a partial mapping from
  operation hashes $\OPH$:
  \begin{align*}
    \PENDING &= \TOPH \partialto \TOP\times \TTIME \times \TSTATUS \\
    \TSTATUS&= (\STATUSPENDING + \STATUSINCLUDED\ \TTIME + \STATUSTIMEOUT)
  \end{align*}

  A pool $\PENDING$ is \emph{well-formed} if, for all $\OPH$, if $\PENDING (\OPH) =
  \Angle{\OP,\TIME, \STATUSINCLUDED\ \TIME'}$, 
  then $\TIME' > \TIME$.
\end{definition}

@@

\begin{definition}%[Blockchain system]
A \emph{blockchain configuration} is a pair
$ \NODE \| \BLOCKCHAIN$ where
\fxnote[inline]{Should the model contain multiple nodes?}
\begin{enumerate}
\item $\NODE = [\EXPRS, \ACCOUNTS, \OPERATIONS]$ is the state of a node, and
\item $\BLOCKCHAIN = [\PENDING, \MANAGERS, \CONTRACTORS,
  \TIME]$ is the state of a blockchain  
  such that
  \begin{itemize}
  \item
    $\forall c \in \ACCOUNTS \implies \exists k\in \MANAGERS, k.\PUK =
    c.\PUK$;
  \end{itemize}
\end{enumerate}
\end{definition}

\section{Transitions}
\label{sec:transitions}

We distinguish several kinds of transitions depending on their scope:
\begin{enumerate}
\item $\ExprTrans$ single-step evaluation of expressions in the node state,
\item $\NodeTrans$ internal transitions of a node,
\item $\BlockTrans$ transitions of the blockchain state,
\item $\SystemTrans$ blockchain system transitions.
\end{enumerate}

Evaluation of expressions is standard call-by-value lambda calculus defined
using evaluation contexts $\EC{}$.
\begin{mathpar}
  \inferrule{}{
    \EC{(\lambda x.\EXPR)\VAL} \ExprTrans \EC{\EXPR[\VAL/x]}
  }

  \inferrule{}{
    \EC{\FST (\VAL_1, \VAL_2)} \ExprTrans \EC{\VAL_1}
  }

  \inferrule{}{
    \EC{\SND (\VAL_1, \VAL_2)} \ExprTrans \EC{\VAL_2}
  }

  \inferrule{
    \TRY \notin \EC[F]{}
  }{
    \EC{\TRY\ {\EC[F]{\RAISE\ \VAL}}\ \EXCEPT\ \EXPR} \ExprTrans
    \EC{\EXPR\ \VAL}
  }

  \inferrule{}{
    \EC{\MATCH\ \NIL\ \WITH\ \NIL\to\EXPR \mid \dots}
    \ExprTrans
    \EC{\EXPR}
  }

  \inferrule{}{
    \EC{\MATCH\ (\CONS\ \VAL_1\ \VAL_2)\ \WITH\ (\CONS\ x_1\ x_2)\to\EXPR \mid \dots}
    \ExprTrans
    \EC{\EXPR[\VAL_1,\VAL_2/x_1, x_2]}
  }

  \inferrule{}{
    \EC{\MATCH\ \NONE\ \WITH\ \NONE\to\EXPR \mid \dots}
    \ExprTrans
    \EC{\EXPR}
  }

  \inferrule{}{
    \EC{\MATCH\ (\SOME\ \VAL)\ \WITH\ (\SOME\ x)\to\EXPR \mid \dots}
    \ExprTrans
    \EC{\EXPR[\VAL/x]}
  }

  \inferrule{}{
    \EC{\MATCH\ (\LEFT\ \VAL)\ \WITH\ (\LEFT\ x)\to\EXPR \mid \dots}
    \ExprTrans
    \EC{\EXPR[\VAL/x]}
  }

  \inferrule{}{
    \EC{\MATCH\ (\RIGHT\ \VAL)\ \WITH\ (\RIGHT\ x)\to\EXPR \mid \dots}
    \ExprTrans
    \EC{\EXPR[\VAL/x]}
  }
\end{mathpar}

The internal transitions of a node are just evaluation of expressions.
\begin{mathpar}
  \inferrule[Node-Eval]
  {
    \EXPR \ExprTrans \EXPR'
  }{
    [\EC\EXPR :: \EXPRS, \ACCOUNTS, \OPERATIONS] \NodeTrans{}
    [\EC{\EXPR'} :: \EXPRS, \ACCOUNTS, \OPERATIONS]
  }
\end{mathpar}
% \begin{mathpar}
%   \inferrule[Node-Op]
%     {
%     \Angle{\PAK,\OP.\PUK} \in \ACCOUNTS
%   }{
%     [\EC\OP :: \EXPRS, \ACCOUNTS, \OPERATIONS] \NodeTrans{}
%     [\EXPRS, \ACCOUNTS, (\EC{},\OP):: \OPERATIONS]
%   }
% \end{mathpar}

\begin{mathpar}
  \inferrule[Node-Inject-NEW]{
    \Angle{\PAK,\OP.\PUK} \in \ACCOUNTS \\
    \CHECKBAL (\MANAGERS, \PUK, \NTEZ, \MTEZ) \\
    \CHECKCOU (\MANAGERS, \PUK) \\
    \CHECKPUB (\MANAGERS, \PUK') \\
    \CHECKGAS (\PUK, \PUK', \NTEZ, \MTEZ) \\
    \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
    \OPH = \GENERATEOPH (\PUK, \PUK', \NTEZ, \MTEZ, \TIME)
  }{
    { [\EC\OP :: \EXPRS, \ACCOUNTS] \|
      [\PENDING, \MANAGERS, \CONTRACTORS, \TIME] } \SystemTrans \\
   { [\EC{\OPH}  :: \EXPRS, \ACCOUNTS] \|
     [ \OPH \mapsto \Angle{\OP, \TIME, \STATUSPENDING}
     ::\PENDING,
     \UPDATECOU(\MANAGERS, \PUK, \TRUE),
     \CONTRACTORS,
     \TIME]
   }
  }
\end{mathpar}

@@

\begin{mathpar}
  \inferrule{
    \CHECKBAL (\MANAGERS, \PUK, \NTEZ, \MTEZ) \\
    \CHECKCOU (\MANAGERS, \PUK) \\
    \CHECKPUB (\MANAGERS, \PUK') \\
    \CHECKGAS (\PUK, \PUK', \NTEZ, \MTEZ) \\
    \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
    \OPH = \GENERATEOPH (\PUK, \PUK', \NTEZ, \MTEZ, \TIME)
  }{
   { [\ACCOUNTS, \OP :: \OPERATIONS] \| [\PENDING,
     \ACCEPTED, \MANAGERS, \CONTRACTORS, \TIME] } \SystemTrans \\
   { [\EXPRS, \ACCOUNTS, \OPERATIONS] \|
     [ \Angle{\OP,\OPH, \TIME}
     ::\PENDING,
     \ACCEPTED,
     \UPDATECOU(\MANAGERS, \PUK, \TRUE),
     \CONTRACTORS,
     \TIME]
   }
  }
\end{mathpar}
~\\
~\\
\fi
Rule 1 [rejected of counter]:
\begin{mathpar}
  \inferrule{
    \NEG\ \CHECKCOU (\BLOCKCHAIN.\MANAGERS, \OP.\PUK) \\
    % \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
  }{
    { [\EC\OP  :: \EXPRS, \ACCOUNTS] \|
      \BLOCKCHAIN
      % [\PENDING, \MANAGERS, \CONTRACTORS, \TIME]
    }
    \SystemTrans
    { [\EC{\RAISE\ \ERRCOUNT} :: \EXPRS, \ACCOUNTS] \|
      \BLOCKCHAIN
      % [ \PENDING, \MANAGERS, \CONTRACTORS, \TIME]
   }
  }
\end{mathpar}
~\\
~\\
Rule 2 [rejected of balance]:
\begin{mathpar}
  \inferrule{
    \NEG\ \CHECKBAL (\BLOCKCHAIN.\MANAGERS, \OP.\PUK, \OP.\NTEZ, \OP.\MTEZ) \\
    % \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
  }{
    { [\EC\OP :: \EXPRS, \ACCOUNTS] \|
      \BLOCKCHAIN
      % [\PENDING, \MANAGERS, \CONTRACTORS, \TIME]
    } \SystemTrans
    { [\EC{\RAISE\ \ERRBAL} :: \EXPRS, \ACCOUNTS] \|
      \BLOCKCHAIN
      % [ \PENDING, \MANAGERS, \CONTRACTORS, \TIME]
   }
  }
\end{mathpar}
~\\
~\\
Rule 3 [rejected of public key]:
\begin{mathpar}
  \inferrule{
    \NEG\ \CHECKPUB (\BLOCKCHAIN.\MANAGERS, \OP.\ADDR) \\
    % \OP = \TRANSFER\NTEZ\PUK{\ADDR}\MTEZ \\
  }{
    { [\EC\OP :: \EXPRS, \ACCOUNTS] \|
      \BLOCKCHAIN
      % [\PENDING, \MANAGERS, \CONTRACTORS, \TIME]
    } \SystemTrans
   { [\EC{\RAISE\ \ERRPUK} :: \EXPRS, \ACCOUNTS] \|
     \BLOCKCHAIN
     % [ \PENDING, \MANAGERS, \CONTRACTORS, \TIME]
   }
  }
\end{mathpar}
~\\
~\\
Rule 4 [rejected of fee]:\todo{Q: is this a local test? move to node transitions?}
\begin{mathpar}
  \inferrule{
    \NEG\ \CHECKGAS (\PUK, \ADDR, \NTEZ, \MTEZ) \\
    \OP = \TRANSFER\NTEZ\PUK{\ADDR}\MTEZ \\
  }{
    { [\EC\OP :: \EXPRS, \ACCOUNTS] \|
      \BLOCKCHAIN
      % [\PENDING, \MANAGERS, \CONTRACTORS, \TIME]
    } \SystemTrans
    { [\EC{\RAISE\ \ERRFEE} :: \EXPRS, \ACCOUNTS] \|
      \BLOCKCHAIN
      % [ \PENDING, \MANAGERS, \CONTRACTORS, \TIME]
   }
  }
\end{mathpar}
\iffalse 
Rule 6 [included]:
\begin{mathpar}
  \inferrule[Block-Accept]{
    \OP = \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
    \TIME'-\TIME < 60
  }{
    {[\Angle{\OP,\OPH, \TIME} :: \PENDING, \ACCEPTED, \MANAGERS,
      \CONTRACTORS, \TIME']}
    \BlockTrans  \\
    {[\PENDING, \Angle{\OP,\OPH, \TIME,\TIME'} :: \ACCEPTED,
      \UPDATESUCC(\MANAGERS, \PUK, \PUK', \NTEZ, \MTEZ),
      \CONTRACTORS, \TIME'+1]}
  }
\end{mathpar}
Rule 7 [timeout]: (applies to both, implicit transfers and contract invocations)
\begin{mathpar}
  \inferrule[Block-Timeout]{
    \OP = \TRANSFER[\STRING]\NTEZ\PUK\ADDR\MTEZ \\
    \TIME'-\TIME \ge 60
  }{
    {[\Angle{\OP,\OPH, \TIME} :: \PENDING, \ACCEPTED, \MANAGERS,
      \CONTRACTORS, \TIME']}
    \BlockTrans
    {[\PENDING, \ACCEPTED, \UPDATECOU(\MANAGERS, \PUK, \FALSE),
      \CONTRACTORS, \TIME']}
  }
\end{mathpar}
\fi
\subsection{Smart Contracts}
A. Originate
~\\
~\\
Rule 1 [injected]:
